### 2023-08-14 chapters/background.tex
Updates

\subsection{Information entropy}
First characterised by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$. Shannon's definition follows from three fundamental axioms of information theory. Let $I_X(x)$ be the information carried by a specific outcome $x$ of sampling the random variable $X$ with probability mass function $p_X(x)$. Then, these axioms are as follows:
\begin{enumerate}[label=(\Roman*)]
    \item The more likely an outcome is, the less information it carries; $I_X(x)$ is a monotonically decreasing function in $p_X(x)$.
    \item Outcomes that are certain to happen carry no information; $I_X(1) = 0$.
    \item The total information carried by independent outcomes is the sum of the information carried by each outcome; $I(x_i, x_j) = I(x_i) + I(x_j)$ if $x_i$ and $x_j$ are independent. 
\end{enumerate}

@@

@@

>>>>

\subsection{Information entropy}
First characterized by Claude Shannon in 1948 \cite{shannon_mathematical_1948}, information entropy, $H(X)$, is a measure of the average amount of information contained in a discrete random variable $X$. Shannon's definition follows from three fundamental axioms of information theory. Let $I_X(x)$ be the information carried by a specific outcome $x$ of sampling the random variable $X$ with probability mass function $p_X(x)$. Then, these axioms are as follows:
\begin{enumerate}[label=(\Roman*)]
    \item The more likely an outcome is, the less information it carries; $I_X(x)$ is a monotonically decreasing function in $p_X(x)$.
    \item Outcomes that are certain to happen carry no information; $I_X(1) = 0$.
    \item The joint information carried by independent outcomes is the sum of the information carried by each outcome; $I(x_i, x_j) = I(x_i) + I(x_j)$ if $x_i$ and $x_j$ are independent. 
\end{enumerate}

@@

\iffalse

@@

\fi

### 2023-08-14 chapters/conclusion.tex
Updates

A conclusion with some abbreviations and citations, \gls{VAE} \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}

>>>>

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation. 
However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.

### 2023-08-14 chapters/discussion.tex
Add benchmarking paper
Add todo to discussion
Updates

%!TEX root = ../thesis.tex
\chapter[discussion]{Discussion}\label{chp:discussion}

\section{}

>>>>

%!TEX root = ../thesis.tex

\chapter[discussion]{Discussion}\label{chp:discussion}

% ~5 pages

\section{\cref{chp:paper-hierarchical} revisited: \dots}

\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other). Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}

\section{\cref{chp:paper-review} revisited: \dots}

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}

### 2023-08-14 chapters/introduction.tex
Minor revisions
Updates

The research done during this project covers a varied set of topics. One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. Another group of studies are concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}. A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.

This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and the main contributions of the thesis.

\subsection{Motivation: Speech recognition}

\subsection{Motivation: Automated medical coding}

\subsection{Challenges in uncertainty estimation}

>>>>

\section{Machine learning and societal risk}
\todo[inline]{Discuss the role of uncertainty aware models in the context of risk.}
\cite{europeancommission_briefing_2021}

\section{Motivation: Speech recognition}

\section{Motivation: Automated medical coding}

\section{Scope of the thesis}

The complete set of research published as part of this project covers a varied set of topics. One group of the produced studies is concerned with generative latent variable models and their applications to uncertainty estimation and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. Another group of studies are concerned with self-supervised speech representation learning and automatic speech recognition \cite{borgholt_scaling_2021,borgholt_we_2021,mohamed_selfsupervised_2022,borgholt_brief_2022}. A final group deals with applications within the medical domain including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.

This first part of the thesis ties together these different studies by providing a high-level overview of the research topic and summarizing the main contributions of the thesis.

### 2023-08-14 chapters/main-contributions.tex
Updates

\chapter[main contributions]{Main contributions}\label{chp:main-contributions}

As previously mentioned, the chapters of the thesis are self-contained studies and therefore detail their own contributions. 
They are however written without consideration to the other chapters. 
For that reason, this chapter will provide a high-level overview of the organization of the thesis and describe its main
contributions in relation to the overall research topic.

\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}

\contribitem{part:background}
This part of the thesis provides the necessary background for the rest of the thesis.

\contribitem{part:unsupervised-learning}

\contribitem{chp:paper-hierarchical}
This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical latent variable models. The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions. To solve this, the proposed method computes a likelihood-ratio score for OOD detection that requires data to be in-distribution across all feature-levels. The method is computationally efficient, scalable, fully unsupervised and achieves state-of-the-art performance on several image-based OOD detection benchmarks.

\contribitem{chp:paper-benchmarking} This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}. The model is benchmarked against other latent variable models and autoregressive models for speech. It is shown that a hierarchy of latent variables improves the likelihood of the model. Finally, the latent space of the models are analyzed in terms of phonetic content.

@@

>>>>

\chapter[main contributions]{Main contributions}\label{chp:main-contributions}
\newcommand{\contribitem}[1]{{\vspace{1em}\noindent\scshape\bfseries{\color{dtured}\cref{#1}}\\\ucnameref{#1}\\}}

The chapters of the thesis are self-contained studies and therefore detail their own contributions. 
They are however written without consideration to the other chapters. 
For that reason, this chapter will provide a high-level overview of the organization of the thesis and describe its main
contributions in relation to the overall research topic.

\contribitem{part:background}
This is the part of the thesis you are currently reading. \Cref{chp:introduction} provided a general introduction to the thesis and gave motivating examples for speech recognition and assistance in medical diagnostics. \Cref{chp:technical-background} provided the necessary technical background for the rest of the thesis by introducing uncertainty as a concept in the context of information theory and probability theory and providing a brief overview of generative latent variable models. 
\lesstodo[inline]{Update the contributions of the technical background section accordingly.}

\contribitem{part:unsupervised-uncertainty-estimation}
This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. The first two papers are concerned with out-of-distribution detection using generative models. The first paper proposes a new method for out-of-distribution detection using hierarchical latent variable models. The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. The third paper proposes a new hierarchical latent variable model for speech

\contribitem{chp:paper-hierarchical}
This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical latent variable models. The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions. To solve this, the proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. The proposed method is computationally efficient, fully unsupervised, and performs well on several image-based out-of-distribution detection benchmarks.
Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.

\contribitem{chp:paper-modelagnostic}
This paper builds on the previous paper and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. The method is based on combining a classical parametric test with the recently introduced typicality test using Fisher's method. We show that this leads to a more accurate out-of-distribution test. We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical out-of-distribution detection, especially in high-risk applications such as in medical devices or services.

\contribitem{chp:paper-benchmarking}
This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}. The model is benchmarked against other latent variable models and autoregressive models for speech abnd we show that using a hierarchy of latent variables improves the likelihood of the model. Finally, we show that the learned latent space captures information relevant for speech recognition such as phonetic content.

\contribitem{part:self-supervised-speech-representation-learning}
This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \cite{mohamed_selfsupervised_2022}. 
Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic methods such as those investigated in \cref{part:unsupervised-uncertainty-estimation}. 
In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning as an alternative paradigm to generative models for learning representations for unsupervised uncertainty estimation. We shall return to this in the discussion of \cref{chp:paper-review}.

\todo[options]{Update this paragraph according to the discussion of self-supervised learning for uncertainty estimation}
% \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.

@@

An early version of this review paper \cite{borgholt_brief_2022} included a review of generative latent variable models for speech representation learning and a comparison between the two modelling paradigms. Due to the superior performance of self-supervised methods and a desire to focus on self-supervised methods, the paper in \cref{chp:paper-review} does not focus on generative latent variable models. The early version is included in \cref{app:paper-brief} for reference.

\todo[inline]{Should we include the brief review paper as a chapter instead of the long review paper?}

\contribitem{part:medical-applications}
This part contains studies on machine learning methods applied for tasks in a medical setting. While uncertainty estimation is not a central theme to any of the two papers, the first paper performs a substantial evaluation of the explainability of the proposed model via an occlusion analysis on the text input. Later, in the discussion, we shall further consider uncertainty estimation in relation to these two papers.
\lesstodo[inline]{Revise and add more details to this section.}

Prior to this thesis, but relevant to mention in this section, we published work on question tracking for calls to emergency services using a multimodal approach that combines automatic transcripts with the raw audio \cite{havtorn_multiqt_2020}.

\contribitem{chp:paper-retrospective}

\contribitem{chp:paper-automated}
